Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-12 _manuscript/_tex/index.tex
revamp toy example

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concepts of diagnosing
unexpected outcomes and general data quality checks.
Section~\ref{sec-plan} introduces the concept of analysis validation
checks, illustrated with a toy example based on fitness step count data.
Section~\ref{sec-method} describes the procedure that quantifies how
analysis validation checks combined using logical operators can predict
unexpected outcomes in an analysis. Section~\ref{sec-pm10-mortality}
applies this procedure to a larger example that estimates the effect of
particulate matter air pollution on daily mortality.
Section~\ref{sec-discussion} discusses a few key considerations and
Section~\ref{sec-conclusion} concludes the paper.

@@

Consider a 30-day step count experiment in public health. Subjects are
instructed to walk at least 8,000 steps each day, with an expected
average of 9,000 steps, tracked by a step counter app. After 30 days, we
review the data and examine the number of steps taken each day. With
data of this nature, we may expect there to be occasional ``low'' days
due to factors such as forgetting to wear their watch or unfavorable
weather conditions limiting outdoor activities. We may also expect
``high'' days recorded after an outdoor hike or intense workout. Given
the requirements of the study, we expect that for a given subject there
would be no more than five days with step counts falling below the
8,000-step threshold (``invalid days''). If there were more than five
invalid days, the data from that subject would not be usable for the
study.

In this scenario, the outcome is that the number of days with a step
count below 8,000 and the expectation is that this outcome would take a
value no more than 5. If the expectation is violated, we can consider
the potential configurations of the data that could lead to such an
anomaly. To diagnose potential reasons why this outcome expectation
might fail, we can establish a few analysis validation checks in
anticipation of seeing the data. For example, if the average step count
is too low, this may suggest that the subject is incapable of taking the
required number of steps, potentially leading to the unexpected outcome.
Similarly, we can also check the quantile of the step count, if more
than a third of the days fall below 8,000, this could indicate an excess
of low-count days. Additionally, we may may expect the standard
deviation of the step count not to be overly large. These considerations
yield the following three analysis validation checks:

\begin{itemize}
\tightlist
\item
  check 1: the check fails if the mean step count is below 8,200
\item
  check 2: the check fails if the 30th percentile of the observed step
  counts is below 8,200
\item
  check 3: the check fails if the standard deviation of the observed
  step counts exceeds 2,500.
\end{itemize}

@@

\caption{\label{fig-step-count}Number of days with fewer than 8,000
steps across 300 simulated 30-day periods. The orange bars indicate
instances where the count exceeds five days, representing an unexpected
outcome in this scenario.}

@@

To simulate this data, three normal distributions are used for the daily
step counts: \(\mathcal{N}(4000, 200)\) for low days,
\(\mathcal{N}(12000, 200)\) for high days, and
\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
days can be simulated from a Poisson distribution with \(\lambda = 4\).
Figure~\ref{fig-step-count} displays the number of days with fewer than
8,000 steps across 300 simulated 30-day periods.

@@

To incorporate the effect of multiple checks on the outcome, a logic
regression model \citep{ruczinski_logic_2003} is fitted to the analysis
validation checks using the indicator of an unexpected result as the
outcome in the model. Originally developed for SNP microarray data,
logic regression constructs Boolean combinations of binary variables, in
a tree structure, to predict both binary and numerical outcomes.
Compared to other tree-based methods for binary-binary prediction, the
Boolean combinations from the logic regression model produce a tree
structure that can be directly interpreted as the possible combination
of checks leading to an unexpected outcome, without the need to invert
the tree as required in classic tree-based recursive partitioning
methods. The logic regression model is then used to predict the analysis
result based on the values of checks, and the prediction is compared to
the actual analysis result in order to calculate the precision and
recall of the checks.

@@

Returning to the step count example introduced in Section~\ref{sec-toy},
the logic regression model is fitted to the three analysis checks
described previously to generate the prediction of the unexpected
outcome, which is whether the number of days with fewer than 8,000 steps
is greater than 5. The predictions from the logic regression model can
be compared with the simulated true outcome for calculating the
precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
best-fitting logic regression model, which is a combination of check 1
(``mean step count is below 8,200'') and check 3 (``standard deviation
exceeds 2,500'') with an OR operator. In other words, if either check 1
or check 3 is true, then we would predict an unexpected outcome in the
analysis (i.e.~the number of days with fewer than 8,000 steps is greater
than 5).

@@

\caption{\label{fig-logic-reg}Logic regression model fitted to the three
unit tests (check 1, check 2, check 3). The model suggests using an OR
rule to combine check 1 and check 3 to predict the outcome expectation.}

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual checks and the combined check rule
(check 1 OR check 3) from the logic regression. We also include the
metric calculated from fitting a regression tree model to the data to
compare the performance of the logic regression model. The harmonic and
arithmetic means are included to combine the three measures. The results
show that the two checks produced by the logic regression can accurately
predict 83.6\% cases of all \emph{actual unexpected results} in the
simulation data. Furthermore, 82.1\% of all \emph{predicted unexpected
results} were in fact observed to be unexpected.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.2468}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1299}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0909}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1688}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1429}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1039}}@{}}

\caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
parsimony (independence) metrics for each individual check and for the
combined check rule (check 1 OR check 3) derived from the logic
regression model. The harmonic and arithmetic means of the three metrics
are included to evaluate the quality of the checks in diagnosing
unexpected step counts (more than five days with fewer than 8,000
steps).}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
tests
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
independence
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
checks
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1 & 0.482 & 0.964 & 1.000 & 0.730 & 0.815 & check 1 \\
2 & 0.214 & 1.000 & 1.000 & 0.450 & 0.738 & check 2 \\
3 & 0.589 & 0.805 & 1.000 & 0.762 & 0.798 & check 3 \\
check 1 OR check 3 & 0.821 & 0.836 & 0.999 & 0.879 & 0.886 & NA \\
regression tree & 0.821 & 0.836 & 1.000 & 0.879 & 0.886 & NA \\

@@

For comparison, the regression tree produces a similar prediction to the
logic regression, by first splitting on check 1 and then splitting on
check 3, and results in the same accuracy and overall score as the logic
regression model. However, we argue that the logic regression tree shown
in Figure~\ref{fig-logic-reg} is more interpretable for our purposes
because it provides a direct representation of which combinations of
analysis checks lead to unexpected outcomes. The logic regression tree
is also directly comparable to other diagnostic techniques, which we
discuss further in Section~\ref{sec-discussion}.

@@

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
check
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
indep.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
check 1: sample size \textless= 200 & 0.087 & 0.215 & 1 & 0.175 &
0.434 \\
check 2: cor(m, PM10) \textless{} -0.03 & 0.988 & 0.610 & 1 & 0.822 &
0.866 \\
check 3: cor(m, tmp) \textgreater{} -0.35 & 0.392 & 0.581 & 1 & 0.569 &
0.658 \\
check 4: mortality\_outlier & 0.649 & 0.641 & 1 & 0.732 & 0.763 \\
(not check 1 AND check 2) AND (check 3 OR check 4) & 0.760 & 0.880 & 1 &
0.869 & 0.880 \\

@@

\section{Conclusion}\label{sec-conclusion}

TODO

>>>>

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concepts of diagnosing
unexpected outcomes and general data quality checks.
Section~\ref{sec-plan} introduces the concept of analysis validation
checks, illustrated with a toy example based on fitness step count data.
Section~\ref{sec-method} describes the procedure that quantifies how
analysis validation checks combined using logical operators can predict
unexpected outcomes in an analysis. Section~\ref{sec-pm10-mortality}
applies this procedure to a larger example that estimates the effect of
particulate matter air pollution on daily mortality.
Section~\ref{sec-discussion} summarises teh paper and discusses a few
key considerations.

@@

Consider a 30-day step count experiment in public health. Subjects are
instructed to walk at least 8,000 steps each day, with an expected
average of 9,000 steps, tracked by a step counter app. After 30 days, we
review the data and examine the number of steps taken each day. With
data of this nature, we may expect there to be occasional ``low'' days
due to factors such as forgetting to wear their watch or unfavorable
weather conditions limiting outdoor activities. We may also expect
``high'' days recorded after an outdoor hike or intense workout. Given
the requirements of the study, we expect that for a given subject the
average step count would be between \([8500, 9500]\).

To diagnose potential reasons why this outcome expectation might fail,
we can establish a few analysis validation checks in anticipation of
seeing the data. For example, we can check the quantile of the step
count, if more than a third of the days fall below 8,000, or more than
two-thirds exceed 10,000 steps, this could indicate an excess of
low-count or high-count days. Similarly, we may may expect the standard
deviation of the step count not to be overly large. These considerations
yield the following three analysis validation checks that \emph{fails}
when:

\begin{itemize}
\tightlist
\item
  check 1: the 60\% quantile of the observed step counts is greater than
  10,000
\item
  check 2: the 40\% quantile of the observed step counts is less than
  8,000, and
\item
  check 3: the standard deviation of the observed step counts exceeds
  2,500.
\end{itemize}

@@

\caption{\label{fig-step-count}Average step count across 300 simulated
30-day periods. The orange points indicate instances where the average
step count fails outside the {[}8,500, 9,500{]} interval, representing
an unexpected outcome in this scenario.}

@@

To simulate this data, three normal distributions are used for the daily
step counts: \(\mathcal{N}(6000, 200)\) for low days,
\(\mathcal{N}(12000, 200)\) for high days, and
\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
days can be simulated from a Poisson distribution with \(\lambda = 8\).
Figure~\ref{fig-step-count} displays average step count across 300
simulated 30-day periods.

@@

To incorporate the effect of multiple checks on the outcome, a logic
regression model \citep{ruczinski_logic_2003} is fitted to the analysis
validation checks. Originally developed for SNP microarray data, logic
regression constructs Boolean combinations of binary variables,
\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict both binary and
numerical outcomes. In our case,

\begin{itemize}
\tightlist
\item
  the outcome, \(Y\), is a binary variable indicating whether the result
  is \emph{unexpected} (1) or \emph{expected} (0),
\item
  \(X_1, X_2, \cdots, X_n\) represent the binary analysis validation
  checks, where each check is labeled as 1 if it fails and 0 if it
  passes.
\end{itemize}

For classification problems, the logic regression model is reduced to
the following form:

\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)\] The logical operations of
the binary variables \(X_1, X_2, \cdots, X_n\), including AND
(\(\land\)), OR (\(\lor\)), and NOT (\(\neg\)) and an example of such
logic combination can be \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\).
Simulated annealing is used to explore the search space by minimizing
the mis-classification error. The following 6 moves are permitted during
the search (as detailed in Figure 2 of \citet{ruczinski_logic_2003}): 1)
replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4)
pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.

Compared to other tree-based methods for binary-binary prediction, the
Boolean combinations from the logic regression model produce a tree
structure that can be directly interpreted as the possible combination
of checks leading to an unexpected outcome, without the need to invert
the tree as required in classic tree-based recursive partitioning
methods. The logic regression model is then used to predict the analysis
result based on the values of checks, and the prediction is compared to
the actual analysis result in order to calculate the precision and
recall of the checks.

@@

Returning to the step count example introduced in Section~\ref{sec-toy},
the logic regression model is fitted to the three analysis checks
described previously to generate the prediction of the unexpected
outcome, whether the average number of step falls within the {[}8,500,
9,500{]} interval. The predictions from the logic regression model can
be compared with the simulated true outcome for calculating the
precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
best-fitting logic regression model as

(quantile(step, 0.6) \textgreater{} 10,000 OR quantile(step, 0.4)
\textless{} 8,000) AND sd(step) \textless{} 2,500

In other words, we would predict an unexpected outcome in the analysis
if the standard deviation of the step count is not too large (2,500) and
either the 60\% quantile of the step count exceeds 10,000 or the 40\%
quantile of the step count falls below 8,000.

@@

\caption{\label{fig-logic-reg}Logic regression model fitted to the three
checks. The model suggests the rule (quantile(step, 0.6) \textgreater{}
10,000 OR quantile(step, 0.4) \textless{} 8,000) AND sd(step)
\textless{} 2,500}

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual checks and the check rule found by
the logic regression. The harmonic and arithmetic means are included to
combine the three measures. The results show that the two checks
produced by the logic regression can accurately predict 86.1\% cases of
all \emph{actual unexpected results} in the simulation data.
Furthermore, 43.1\% of all \emph{predicted unexpected results} were in
fact observed to be unexpected.

\caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
parsimony (independence) metrics for each individual check and the logic
regression check rule. The harmonic and arithmetic means of the three
metrics are included to evaluate the quality of the checks in diagnosing
unexpected step counts (more than five days with fewer than 8,000
steps).}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
checks
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
indep.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
check 1: q(step, 0.6) \textgreater{} 10000 & 0.319 & 0.575 & 1.000 &
0.511 & 0.631 \\
check 2: q(step, 0.4) \textless{} 8000 & 0.264 & 0.613 & 1.000 & 0.467 &
0.626 \\
check 3: sd(step) \textgreater{} 2500 & 0.153 & 0.289 & 1.000 & 0.273 &
0.481 \\
logic regression: (check 1 OR check 2) AND check 3 & 0.431 & 0.861 &
0.998 & 0.669 & 0.763 \\
regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774 \\

@@

For comparison, the regression tree produces a similar prediction to the
logic regression, however, we argue that the logic regression tree shown
in Figure~\ref{fig-logic-reg} is more interpretable for our purposes
because it provides a direct representation of which combinations of
analysis checks lead to unexpected outcomes. The logic regression tree
is also directly comparable to other diagnostic techniques, which we
discuss further in Section~\ref{sec-discussion}.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.6106}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0885}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0619}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0619}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0796}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0973}}@{}}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
check
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
indep.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
check 1: sample size \textless= 200 & 0.087 & 0.215 & 1 & 0.175 &
0.434 \\
check 2: cor(m, PM10) \textless{} -0.03 & 0.988 & 0.610 & 1 & 0.822 &
0.866 \\
check 3: cor(m, tmp) \textgreater{} -0.35 & 0.392 & 0.581 & 1 & 0.569 &
0.658 \\
check 4: mortality\_outlier & 0.649 & 0.641 & 1 & 0.732 & 0.763 \\
logic regression: (not check 1 AND check 2) AND (check 3 OR check 4) &
0.760 & 0.880 & 1 & 0.869 & 0.880 \\

@@
